On the effects of partying on next-day sports performance

### 2017-06-01 writing/draft.tex
Tightened up first rough draft with Eric.

\section{Introduction}
It's not unreasonable to think that game performance can be affected 
(in part) by what takes place off the court. 
Anecdotally, it has been reported that professional sports teams party on 
the road, sometimes to the detriment of the game.
Within the NBA, J.R. Smith is infamous for citing his improvement in performance due to a
lack of nightlife when moving to Cleveland from New York.\cite{price,ley}
Matt Barnes recently complained that ``there's no nightlife in Utah'' when 
asked about playing Jazz in the second round of NBA playoffs in 2017.\cite{cestone}
A few interviews with former professional NBA athletes revealed that location
matters more than day of the week, i.e. parties can happen any day of the week,
but they are more likely to occur in select cities.

@@

\section{Materials and Methods}
One of the hardest parts of this analysis is that partying is a latent variable.
In order to determine when players partied, one might consider to look at social media data
for direct mentions of partying, either by athletes or witnesses. However, due to
intense public scrutiny that athletes face, there is a strong reluctance to disclose
their personal lives on social media. There are notable exceptions, such
as Oddel Beckham Jr.'s recent trip to Miami.\cite{bleler}
However, these reports are few and far enough between,
and whether they are even reported at all is strongly biased on whether a particular city 
has nightclubs with VIP rooms, etc; picking up statistical evidence would be fruitless in this case.

@@

We then merge this in with our lines data, taking care to do so according to each season-team's last game location such that the feature captures the proper
party effect. In addition, we lag our BLS data by one year for two reasons:
realistically, player expectations of which cities have the most active night-life may not
update in real-time and also doing so allows us to create a betting model, i.e.
we don't rely on any features we can't obtain before making a game-time prediction. 
To this data, we've created features such as 
the number of days since last game, the last game location, and the travel distance.

@@

\subsection{Method: Defining what it means for a city to have nightlife}
We have a free parameter: the time-window which we believe the hangover effect
to last. Suppose we only look toward games which have been played in a party
city at most 24 hours ago. We take care to ignore
teams playing in their own cities, i.e. if a team's last game was played in at home 
we say that they were not exposed to an exogenous shock in nightlife.

@@

In general, we can replicate our analysis and use the same resources to obtain our data,
 but there are also some notable differences.
In the MLB, money-lines
are used instead of point-spreads. So as opposed to predicting
the likelihood of meeting the spread, we predict the probability
a team would win against their opponent as both a function of the book-makers'
odds of winning and our party variable. 
Additionally, our methodology changes ever so slightly: when creating our
continuous measure of nightlife, we interact the variable with weekend. The logic
is that because there are so many more games played in MLB, the weekend effect is more salient.
One of the last major differences between NBA and MLB games is that MLB games
are played in series, so the same match-up is repeated multiple times. This remedies
the problem of an unknown travel schedule we faced with NBA because we are sure 
teams stay overnight in the same location within a series.

@@

\subsubsection{Betting performance}
We also validate our MLB model by seeing how well we can perform against the
market.
\begin{figure}[!h]
  \centering
  \includegraphics{mlb_bets_by_season}
  \caption{Betting performance by season. In each season, The 
    circled point represents the worst out of pocket expense incurred.
    We first train a model on historical data then use said model to predict on ``current year''
    data, plotting profits as a function of time over the ``current'' season. For each relevant
    game, we place a bet of \$100 if the expected value of the bet is positive after accounting f
    or the bookmaker's cut. E.g. in the     early 2016 season,
    at its worst we are out of pocket \$750 dollars. By season end, we have     recouped all of
    our initial investment plus an additional \$11.5k.}
  \label{bettingperf}
\end{figure}

@@

\subsection{Extending the research}
We can also validate where points are lost in MLB in a similar way.
It would be interesting to look at number of shots taken, and number of shots taken by stars.
It's possible there are prolonged effects of partying, and applying some sort of kernel
on the number of hours played since the last game could be interesting. It could also be
that different (types of) teams or players are more or less susceptible to the effects
of night-life; this could be demographic characteristics such as age or marital status.
Future work may consider similar effects in European soccer leagues.

\section{Conclusions}
We do find that visiting a city with active nightlife the day before a game does have 
a statistically significant negative impact on a team's likelihood of meeting the spread in NBA.
The decline in likelihood to meet the spread is possibly caused by a lack of control over the ball,
since we also find that players capture less rebounds per minute of gameplay the day after
visiting a city with active nightlife. We verify that this phenomena persists across sport, and 
to that end look toward major league baseball money-lines. Using these, we back out gambling
house's estimate for the team's probability of winning the game, and after conditioning on this
we find that visiting a city with active nightlife the day before a game does negatively 
impact the team's likelihood of winning the game.
With a causal model established, we realize a profitable betting strategy in each year of testing.

>>>>

\section{Introduction}
It's not unreasonable to think that game performance can be affected 
(in part) by what takes place off the court. 
Anecdotally, it has been reported that professional sports teams party on 
the road, sometimes to the detriment of the game.
Within the NBA, the player J.R. Smith underwent a substantial improvement
in performance metrics after being traded from New York to Cleveland; when an interviewer
asked what contributed to this bump in statistics, he infamously replied
that in New York he would have been out partying, but because there's no nightlife
in Cleveland, he ends up hitting the gym.\cite{price,ley}
Matt Barnes recently complained that ``there's no nightlife in Utah'' when 
asked about playing Jazz in the second round of NBA playoffs in 2017.\cite{cestone}
A few interviews with former professional NBA athletes revealed that location
matters more than day of the week, i.e. parties can happen any day of the week,
but they are more likely to occur in select cities.

@@

\section{Materials and Methods}
One of the hardest parts of this analysis is that partying is a latent variable.
In order to determine when players partied, one may consider looking at social media data
for direct mentions of partying, either by athletes or witnesses. However, due to the
intense public scrutiny that athletes face, there is a strong reluctance to disclose
their personal lives on social media. There are notable exceptions, such
as Oddel Beckham Jr.'s recent trip to Miami.\cite{bleler}
However, these reports are few and far enough between that picking up statistical 
evidence would be fruitless.

@@

We then merge this in with our lines data, taking care to do so according to each season-team's last game location such that the feature captures the proper
party effect. In addition, we lag our BLS data by one year for two reasons:
realistically, player expectations of which cities have the most active night-life may not
update in real-time and also doing so allows us to create a betting model, i.e.
we don't rely on any features we can't obtain before making a game-time prediction. 
To this data, we've added features such as 
the number of days since last game, the last game location, and the travel distance.

@@

\subsection{Method: Defining what it means for a city to have nightlife}
We have a free parameter: the time-window which we believe the hangover effect
to last. We consider games which have been played at most 24 hours ago, and take care to ignore
teams playing in their own city, i.e. if a team's last game was played at home  we say that they were not exposed to an exogenous shock in nightlife.

@@

In general, we can replicate our analysis and use the same resources to obtain our data,
 but there are also some notable differences.
In the MLB, money-lines
are used instead of point-spreads. So as opposed to predicting
the likelihood of meeting the spread, we predict the probability
a team would win as both a function of the book-makers'
odds of winning and our party variable. 
Additionally, our methodology changes ever so slightly: when creating our
continuous measure of nightlife, we interact the variable with weekend. Interviews with athletes suggest that MLB players are more likely to
restrict their partying to weekends.
One of the last major differences between NBA and MLB games is that MLB games
are played in series, so the same match-up is repeated multiple times. This remedies
the problem of an unknown travel schedule we faced with NBA because we are sure 
teams stay overnight in the same location within a series.

@@

As shown above, our continuous measure of nightlife remains statistically significant.

\subsubsection{Betting performance}
We also validate our MLB model by seeing how well we can perform against the
market.
\begin{figure}[!h]
  \centering
  \includegraphics{mlb_bets_by_season}
  \caption{Betting performance by season. In each season, The 
    circled point represents the worst out of pocket expense incurred.
    We first train a model on historical data then use said model to predict on ``current year''
    data, plotting profits as a function of time over the ``current'' season. For each relevant
    game, we place a bet of \$100 if the expected value of the bet is positive after accounting for the bookmaker's cut. E.g. in the     early 2016 season,
    at its worst we are out of pocket \$750 dollars. By season end, we have     recouped all of
    our initial investment plus an additional \$11.5k.}
  \label{bettingperf}
\end{figure}

@@

\subsection{Extending the research}
\paragraph{Within the NBA} 
We're interested to explore what other factors may affect a team's propensity to suffer hangover effects beyond simply visiting a party city; e.g. are teams more likely to go out after
a win or a loss? Do teams who aren't going to make the playoffs party more? Are there demographic
factors which influence partying, e.g. are younger, unmarried players more susceptible to treatment effect? Within a game, it would be interesting to look at 
number of shots taken, and number of shots taken by 
stars as a function of previous day visitation to party cities. We could also investigate 
effects on injury risk and flagrant foul rates.

\paragraph{MLB}
It would also be interesting to examine the particular ways that partying
effects performance statistics similar to how we did with NBA.
We might consider whether batting performance (averages, slugging percentage)
pitching (ERA, WHIP) or general fielding stats are effected by party.

\paragraph{General}
In both sports, it's possible there are prolonged effects of partying, and applying some sort of kernel on the number of hours played since the last game could be interesting. 
Future work may consider similar effects in European soccer leagues.

\section{Conclusions}
We do find that visiting a city with active nightlife the day before a game does have 
a statistically significant negative impact on a team's likelihood of meeting the spread in NBA. This decreased likelihood to meet the spread may be explained 
by the team's defense letting up more points. In turn, we believe that the drop in
defensive performance may be caused by a drop in physical exertion.
We verify that this phenomena persists across sport, and 
to that end look toward major league baseball money-lines. Using these, we back out gambling
house's estimate for the team's probability of winning the game, and after conditioning on this
we find that visiting a city with active nightlife the day before a game does negatively 
impact the team's likelihood of winning.
With a causal model established, we realize a profitable betting strategy in each year of testing.
